% =====================================================================================
% book.tex — Causal Marketing: the whole book.
%
%   xelatex book && biber book && xelatex book && xelatex book
%   (or, correctly: python book/build.py — which regenerates macros.tex first)
%
% Chapters are \include'd, so \includeonly{chapters/geo_lift} produces the pilot chapter
% alone from this same source; book/chapter.tex is the standalone driver that does exactly
% that with its own front matter.
% =====================================================================================
\documentclass[11pt, a4paper, twoside, openright, numbers=noenddot,
               chapterprefix=true, DIV=11, BCOR=8mm]{scrbook}

\input{preamble}
\addbibresource{refs.bib}

% Every number below arrives from an executed notebook. Nothing is retyped.
\input{build/macros}

\title{Causal Marketing}
\subtitle{Classical estimates, Bayesian posteriors, and the euro decision}
\author{Francesco Muia}
\date{\today}

\begin{document}

\frontmatter
\maketitle

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book is about a narrow and expensive question: \emph{did the marketing spend cause the
sales, and was it worth the money?} It is written for the analyst who must answer it with the
data a company actually has --- a weekly sales panel, a campaign flag, a cost line --- and who
must then defend the answer to someone holding the budget.

Each chapter takes one marketing decision and works it twice. First classically: the simplest
correct estimator, a point estimate, an interval whose covariance assumption is stated out loud.
Then Bayesianly: a likelihood, priors that are justified rather than defaulted, a posterior. The
two answers are then laid side by side on the \emph{same} estimand, and the chapter says plainly
where they agree, where they differ, and why. Usually they agree; the interesting chapters are the
ones where they do not, and the book's central claim --- earned, not asserted --- is that when the
Bayesian arm loses, the culprit is a \emph{misspecified likelihood}, not the Bayesian paradigm.
The chapter you are about to read, if you are reading the pilot, is the sharpest instance of that
claim in the book.

Two conventions carry through. Every result is reported as a captioned table or figure and
referred to by number; there is no pasted program output in these pages. And no number in the
prose was typed by a human: each one is a macro injected from the notebook that computed it, so a
figure in the text cannot drift from the figure in the output. A missing number is a compilation
error, which is the property we wanted.

\tableofcontents
\listoffigures
\listoftables

\mainmatter

% ---------------------------------------------------------------------------------------
% Part II · Methods.  (Part I — foundations — and chapters 3–8, 10–13 are drafted in the
% notebooks and land in this file as they are converted; the geo-lift chapter is the pilot
% and fixes the template for all of them.)
% ---------------------------------------------------------------------------------------
% The geo-lift chapter is Chapter 9 of the planned book (BOOK_SPEC.md), and is numbered as
% such: its cross-references, equation numbers and table numbers are the ones it will keep
% when the chapters around it land. Advancing the counters is what makes the pilot a genuine
% chapter of a whole rather than a standalone note that will have to be renumbered later.
\setcounter{part}{1}
\part{Methods}
\setcounter{chapter}{8}

\include{chapters/geo_lift}

\backmatter
\printbibliography[heading=bibintoc, title={Bibliography}]

\end{document}
