% =====================================================================================
% Chapter 1 · Potential outcomes, confounding, and DAGs
%
% Built from notebooks/00_foundations.ipynb (§1–§6, §8, §9). NOT ONE NUMBER IN THIS FILE
% IS TYPED BY HAND: every figure in the prose is a macro from build/macros.tex, generated
% by cmp.report from the executed notebook (CMP_FAST=0, CMP_REAL=1).
%
% STRUCTURE — this chapter owns the book's vocabulary, so it opens differently from the
% uniform method-chapter template. §1.2 ("Setup and notation") is a compact primer: it fixes
% the running example and the symbols (unit, T, Y, x; the estimands; the propensity; the naive
% difference) with a one-line gloss each, and states every identity in plain language. The
% formal machinery — the potential-outcomes definitions, the selection-bias derivation,
% d-separation and the fork/chain/collider rules, the g-formula, the estimator standard errors,
% and the balance diagnostic — lives in the per-chapter Appendix 1.A, each with the worked
% example the body points to. The body keeps the plain statements, the euro results, and the
% pointers; the appendix keeps the derivations. §1.5 APPLIES the graph rules that 1.A defines,
% so the fork/chain/collider material appears once, not twice.
%
% The only literals below are (i) conventional thresholds (the 0.1 SMD bar, the [0.05, 0.95]
% propensity clip, the 90 % interval level), which are conventions stated in the text, and
% (ii) the published NSW experimental benchmark, which is emitted as a macro and flagged, in
% the notebook and here, as cited rather than estimated. Every DGP constant that used to sit
% bare inside a displayed equation is now a named symbol whose value lives in Table 1.3
% (tab:fnd:params) and in the book's global Appendix A.
% =====================================================================================
\chapter{Potential outcomes, confounding, and DAGs}
\label{ch:foundations}

\begin{quote}\small
\emph{A retailer's dashboard says emailed customers spend \eur{\nbZeroObsNaive} more than
un-emailed ones, against a contact cost of \eur{\nbZeroCost}. Send the email to everyone: it
clears its cost. That recommendation is wrong. The true average effect here is only
\eur{\nbZeroTrueAte} --- below the cost --- and the gap is not noise. The confounded estimate
sits \nbZeroObsSeDist{} standard errors from the truth; the randomized one sits
\nbZeroRandSeDist{}. Exactly \eur{\nbZeroSelBias} of the gap is a statement about \emph{who} was
emailed, not about what the email did. Acting on the dashboard burns \eur{\nbZeroPolBurn} across
\nbZeroNCustomers{} customers. This chapter builds the vocabulary that makes that precise ---
potential outcomes, confounding, backdoor paths, colliders. It then climbs the estimator ladder
that repairs the number: a naive difference of means, then regression adjustment, then
inverse-propensity weighting, then the doubly-robust estimator, each with the standard error its
data demands. It closes on LaLonde's job-training data. There the naive comparison says training
destroyed \usd{\nbZeroLlNaive} of earnings; adjustment moves it \usd{\nbZeroLlSwing}, across
zero, to within \usd{\nbZeroLlShortfall} of the randomized benchmark.}
\end{quote}

% =====================================================================================
\section{The decision}
\label{sec:fnd:decision}

A retailer has a customer file and an email programme. Each contact costs \eur{\nbZeroCost}
to print, post and process. The analytics team pulls the obvious number --- average spend
among customers who were emailed, minus average spend among those who were not --- and it
comes to \eur{\nbZeroObsNaive}. The email clears its cost by a comfortable margin. Scale it.

That recommendation is the most expensive sentence in marketing analytics, and it is wrong
for a reason that has nothing to do with the arithmetic. The arithmetic is correct: the
emailed customers really did spend \eur{\nbZeroObsNaive} more. What is wrong is the
\emph{claim} smuggled into the comparison --- that the un-emailed customers tell us what the
emailed ones \emph{would have spent} had no email been sent. They do not, because nobody
flipped a coin. A marketer chose whom to email, and chose the loyal, engaged, recently-active
customers, who spend more anyway.

The chapter's central quantity is therefore not a number but a \emph{missing} number: what
the treated would have done untreated. Everything that follows --- the notation, the graphs,
the four estimators, the diagnostics --- exists to reconstruct it, and to state honestly the
assumptions under which the reconstruction is licensed. \Cref{sec:fnd:euro} prices what
getting it wrong costs, and the figure it arrives at, \eur{\nbZeroPolBurn} burned per
\nbZeroNCustomers{} contacts, is the reason this is not a philosophy chapter.

One convention, used throughout. Spend is booked as \emph{contribution}, with the unit margin
normalised to one, so a euro of incremental spend is a euro of profit set against the
\eur{\nbZeroCost} a contact costs. A lower margin only strengthens the no-send verdict of
\cref{sec:fnd:euro}: less profit per euro of spend pushes the decision further onto the side it
already lands on.

% =====================================================================================
\section{Setup and notation}
\label{sec:fnd:vocab}

This section fixes the running example and the symbols the rest of the book uses. It states
each idea in plain language and gives a one-line gloss; the formal definitions, the derivations,
and the worked examples that go with them live in Appendix~1.A (\emph{Formal machinery}) and are
pointed to as each symbol appears. A reader who already knows potential outcomes, selection bias
and $d$-separation can skim the glosses and go to \cref{sec:fnd:dgm}.

\subsection*{The unit, the treatment, the outcome --- and the file they live in}

Three objects, and everything else is built from them. The \textbf{unit} is the thing a
decision is taken about: here, a customer. The \textbf{treatment} $T$ is the action that could
be taken on it: $T = 1$ if the customer is emailed, $T = 0$ if not. The \textbf{outcome} $Y$
is what the business cares about afterwards: spend, in euros. Each unit also carries
\textbf{features} $x$ --- recency $R$, frequency $F$, monetary value $M$, tenure, engagement
$E$ --- which are properties of the customer and are \emph{not} affected by the email.

\Cref{tab:fnd:sample} is the file itself: eight real rows of the simulated customer base this
chapter uses. It is worth looking at before any algebra, because every abstraction below is a
statement about these columns.

\input{build/tables/nb00_sample_rows}

\subsection*{The notation, at a glance}

Fix a customer $i$. Before anything is done she carries \emph{two} numbers, not one: $Y_i(1)$,
the spend if emailed, and $Y_i(0)$, the spend if not. These are her \textbf{potential outcomes}
\citep{neyman1923, rubin1974}. The email reveals exactly one of them and hides the other forever
--- the customer was emailed, so the world in which she was not is gone. That is the
\emph{fundamental problem of causal inference} \citep{holland1986}, and its most useful reading
is \citet{ding2018}'s: \textbf{causal inference is a missing-data problem}. Half of every
customer's row was never filled in. Appendix~1.A states this formally (consistency,
\cref{eq:fnd:consistency}) and shows it on six real customers whose hidden halves the simulator
knows (\cref{tab:fnd:oracle}).

The symbols, each with the one gloss the body relies on:

\begin{description}[leftmargin=*, style=nextline, itemsep=.2em]
  \item[Unit, treatment, outcome, features.] A \textbf{unit} is a customer; the
    \textbf{treatment} $T$ is $1$ if she is emailed and $0$ if not; the \textbf{outcome} $Y$ is
    her spend in euros; the \textbf{features} $x$ (recency, frequency, monetary value, tenure,
    engagement) are fixed traits the email does not change.
  \item[Individual effect $\tau_i = Y_i(1) - Y_i(0)$.] What the email did to \emph{that}
    customer. Permanently invisible, because one of its two terms is always missing.
  \item[The three averages.] Decisions run on averages of $\tau_i$, and \emph{which} average is
    a commercial choice: the \textbf{ATE} prices \emph{roll it out to everyone?}, the
    \textbf{ATT} audits \emph{what did our past sends earn?}, and the \textbf{CATE} $\tau(x)$
    decides \emph{who next?} They are defined in Appendix~1.A and are not interchangeable ---
    \cref{fig:fnd:tau} shows why.
  \item[Propensity $e(x) = \Prob(T{=}1 \mid X{=}x)$.] The probability a customer with features
    $x$ gets emailed \citep{rosenbaum1983}. It runs the weighting estimator of
    \cref{sec:fnd:classical} and the overlap assumption of \cref{sec:fnd:ident}.
  \item[Naive difference $\hat\tau_{\text{naive}}$.] The dashboard's number: observed average
    spend among the emailed minus among the rest.
\end{description}
%
\begin{equation}
  \hat\tau_{\text{naive}} \;=\; \bar Y_1 - \bar Y_0
    \;=\; \widehat{\E}\big[Y \mid T{=}1\big] - \widehat{\E}\big[Y \mid T{=}0\big].
  \label{eq:fnd:naive}
\end{equation}
%
The naive difference is computable from any customer file and needs no assumptions to
\emph{compute}. Whether it \emph{means} anything causal is the entire question --- and its
standard error, the typical distance from the truth due purely to sampling luck, has no opinion
on that at all (Appendix~1.A gives the Welch form, \cref{eq:fnd:welchse}). A confounded estimate
with a small standard error is not a good estimate; it is a precisely wrong one.

Which average matters is commercial, not terminological. \Cref{fig:fnd:tau} makes the point: the
ATE of \eur{\nbZeroTrueAte} sits \emph{below} the \eur{\nbZeroCost} cost, so an average-level
rule says \emph{email no one} --- and yet \pc{\nbZeroShareAboveCost} of customers individually
clear the cost bar, while \pc{\nbZeroShareNegative} are actively \emph{harmed} by the email. One
number, opposite per-customer instructions. Recovering $\tau(x)$ is the subject of \Cref{ch:up};
a fourth average, the \textbf{LATE}, needs instruments and so waits for \Cref{ch:iv}.

\input{build/floats/nb00_tau_spread}

\subsection*{Selection bias, and the identity that runs the chapter}

The naive difference is not the causal effect. It is the effect \emph{plus} a second term, and
the two come apart in an exact identity that runs the whole chapter:
%
\begin{equation}
  \underbrace{\E[Y \mid T{=}1] - \E[Y \mid T{=}0]}_{\text{the naive difference (observable)}}
  \;=\;
  \underbrace{\E[Y(1) - Y(0) \mid T{=}1]}_{\text{ATT: a causal effect}}
  \;+\;
  \underbrace{\E[Y(0) \mid T{=}1] - \E[Y(0) \mid T{=}0]}_{\text{selection bias}} .
  \label{eq:fnd:decomp}
\end{equation}
%
\textbf{Selection bias} is the \emph{head start}: the gap in baseline spend between the emailed
and the un-emailed that would have existed \emph{even if nobody had been emailed at all}. It
answers ``were these two groups comparable to begin with?'' --- a question about who was chosen,
not about what the email did. The naive difference silently substitutes the controls' baseline,
$\E[Y(0) \mid T{=}0]$, which is observed, for the treated group's own counterfactual baseline,
$\E[Y(0) \mid T{=}1]$, which is not. Appendix~1.A derives \cref{eq:fnd:decomp} line by line and
watches it hold on the six oracle customers of \cref{tab:fnd:oracle}.

On the full file of \nbZeroNCustomers{} customers the identity is exact and damning. The
dashboard's naive difference of \eur{\nbZeroNaiveObs} is an ATT of \eur{\nbZeroAtt} plus
\eur{\nbZeroSelBias} of selection bias. And there is a \emph{second} error hiding inside the
first: the ATT --- the effect on the customers actually emailed --- exceeds the ATE by
\eur{\nbZeroTargetingGap}, because past targeting favoured the very customers on whom the email
works best. So the \eur{\nbZeroObsNaive} the dashboard reports and the \eur{\nbZeroTrueAte} the
business needs are separated by \emph{two} distinct errors: selection bias, and a question about
which estimand was wanted. \Cref{fig:fnd:selgap} shows the first on the full customer file --- the
two baseline distributions, their means, and the gap the naive difference mistakes for an effect.

\input{build/floats/nb00_selection_gap}

\subsection*{Randomization, and what it buys}

Suppose the marketer flips a coin instead of choosing. Then $T$ is independent of
\emph{everything} the customer walks in with --- measured or not, known or not --- and in
particular of $Y(0)$. Independence means the emailed and the un-emailed have the \emph{same}
baseline distribution, so
%
\[
  \E[Y(0) \mid T{=}1] \;=\; \E[Y(0) \mid T{=}0] \;=\; \E[Y(0)],
\]
%
and the selection-bias term in \cref{eq:fnd:decomp} is \textbf{exactly zero}. Not small: zero.
No group has a head start, because who was chosen has nothing to do with who was going to
spend. The naive difference then estimates the ATT, which now equals the ATE, and
\cref{eq:fnd:naive} --- the dashboard's own number --- becomes a valid causal estimate.

That is the whole proof, and it is why an A/B test is worth more than every estimator in this
book. The rest of the book exists for the case where the coin was never flipped.

\subsection*{Graphs, and what to adjust for}

The last piece of vocabulary decides \emph{what to adjust for}. A \textbf{directed acyclic
graph} (DAG) draws each variable as a node and each direct cause as an arrow. It is a claim about
the world, committed to before the data are touched \citep{greenland1999}. Three configurations
exhaust the ways two variables can connect, and conditioning does something different to each
(\cref{fig:fnd:dags}):

\begin{itemize}[leftmargin=*, itemsep=.2em]
  \item \textbf{Fork} ($T \leftarrow L \rightarrow Y$): loyalty drives both the send and the
        spend. Adjust for it, or the association is not causal.
  \item \textbf{Chain} ($T \rightarrow M \rightarrow Y$): the email works by being opened. Do not
        adjust, or you delete part of the effect.
  \item \textbf{Collider} ($T \rightarrow C \leftarrow Y$): responding is caused by both. Do not
        condition, or you manufacture association from nothing.
\end{itemize}

\input{build/floats/nb00_dags}

A \textbf{backdoor path} is confounding drawn explicitly. To \textbf{adjust} for $X$ is to
compare like with like within $X$ and average over the population --- the \emph{g-formula}
(\cref{eq:fnd:gformula}). Appendix~1.A defines all of this; \cref{sec:fnd:collider} applies it
and prices what getting it wrong costs.

\begin{recap}
A unit has two potential outcomes and reveals one (\cref{eq:fnd:consistency}); the missing one
is the problem. Decisions use averages --- ATE, ATT, CATE --- and which one matters
commercially. The naive difference (\cref{eq:fnd:naive}) equals the ATT \emph{plus} selection
bias (\cref{eq:fnd:decomp}), where selection bias is the treated group's head start. A coin
flip makes that term exactly zero. Without a coin flip, we must \emph{adjust}: compare like
with like within strata of $X$ and average over the population
(\cref{eq:fnd:gformula}) --- and the DAG is what tells us which variables belong in $X$.
Everything from here on is a way of doing that, a way of checking it, or a way of showing
what it costs when it fails.
\end{recap}

% =====================================================================================
\section{The data-generating model}
\label{sec:fnd:dgm}

A causal method cannot be graded on the data a business actually has, because there the
counterfactual is missing --- that is the entire problem. But it can be graded, and this book
grades it three ways, in strictly decreasing order of what is known:
%
\begin{enumerate}[leftmargin=*, itemsep=.2em]
  \item \textbf{A simulated world}, where the answer is known \emph{by construction} because we
        planted it. Every estimator can be scored exactly. This section builds three.
  \item \textbf{A within-study benchmark}, where the answer is known \emph{by randomization}:
        a dataset that carries both an experimental arm and an observational one, so an
        observational method can be scored against the experiment. \Cref{sec:fnd:real} sits that
        exam on LaLonde; \Cref{ch:geo} sits it on a real geo experiment.
  \item \textbf{Deployment}, where there is no benchmark at all and never will be. Here a method
        cannot be certified --- only \emph{falsified}, by tests it could have failed and did not
        (placebos, negative controls, overidentification). That is the whole of the diagnostic
        apparatus in \cref{sec:fnd:valid}, and it is one-sided by nature.
\end{enumerate}
%
The three tiers are not interchangeable, and confusing them is how a field talks itself into
believing an unfalsifiable estimate. This chapter uses the first two, and says which is which.

\subsection*{The customer file}

Each of the \nbZeroNCustomers{} customers carries the features $x$ of \cref{tab:fnd:sample},
a baseline spend $\mu_0(x)$ --- what she would spend with no email --- and a planted
per-customer effect $\tau(x)$. Observed spend is
%
\begin{equation}
  Y \;=\; \mu_0(x) \;+\; \tau(x)\,T \;+\; \varepsilon,
  \qquad \varepsilon \sim \mathcal N\!\left(0,\ \sigma_\varepsilon^{2}\right),
  \label{eq:fnd:uplift}
\end{equation}
%
with $T \in \{0,1\}$ the email. Comparing \cref{eq:fnd:uplift} with the definitions of
\cref{sec:fnd:vocab}: $Y(0) = \mu_0(x) + \varepsilon$ and $Y(1) = \mu_0(x) + \tau(x) +
\varepsilon$, so the individual effect $\tau_i$ \emph{is} $\tau(x_i)$, and the ATE of
\cref{eq:fnd:ate} is its average over the file, $\bar\tau = \E[\tau(x)]$. That average is
\eur{\nbZeroTrueAte}, and it is known for one reason only: we planted it. The exact shape of
$\tau(x)$ --- a bump that is positive for persuadable customers and negative for the ``sleeping
dogs'' an email annoys --- is given in Appendix A; \cref{fig:fnd:tau} is its histogram, and
\cref{tab:fnd:sample} shows individual customers sitting at both ends of it.

The simulator is then run \emph{twice}, and only the assignment rule differs:
%
\begin{align}
  \text{randomized:}\quad     & T \sim \text{Bernoulli}(\tfrac12), \label{eq:fnd:rct}\\
  \text{observational:}\quad  & T \sim \text{Bernoulli}\Big(\sigma\big(
      \gamma_E (E - \bar E) + \gamma_R \tilde R + \gamma_F (F - \bar F)\big)\Big),
      \label{eq:fnd:obs}
\end{align}
%
where $\sigma(z) = 1/(1 + e^{-z})$ is the logistic function, which turns any score into a
probability, $\tilde R$ is recency rescaled so that recent is high, and the coefficients
$\gamma$ are positive (their values are in Appendix A). \Cref{eq:fnd:obs} is a caricature of a
real marketing department: engaged, recently-active, frequent buyers are the ones who get
emailed --- and, through $\mu_0(x)$, they are the ones who would have spent more regardless.

Two consequences, and they are the reason the chapter exists. First, the planted average effect
is \eur{\nbZeroTrueAte} in \emph{both} worlds: $\tau(x)$ is a property of the customer, and
nothing in \crefrange{eq:fnd:rct}{eq:fnd:obs} touches it --- only the distribution of $T$
changes. Second, that change alone is enough to move the naive difference of
\cref{eq:fnd:naive} from \eur{\nbZeroRandNaive} to \eur{\nbZeroObsNaive}
(\cref{fig:fnd:confounding}). Same customers, same effect, same estimator; a different rule for
who gets picked, and the answer nearly doubles. \Cref{fig:fnd:assign} shows what those two rules
actually \emph{do}.

\input{build/floats/nb00_assign_mech}
\input{build/floats/nb00_confounding}

\subsection*{The one-confounder world}

The estimator ladder of \cref{sec:fnd:classical} needs a world simple enough that the reader
can hold the entire causal structure in mind, and it gets one. A single \emph{observed}
confounder, loyalty $L$, drives both the send and the spend. A second trait $G$ moves spend but
never touches assignment --- so it is \emph{not} a confounder, and it is in the model
deliberately, as a foil:
%
\begin{align}
  T &\;\sim\; \text{Bernoulli}\big(\sigma(\alpha_L\, L)\big), \label{eq:fnd:dag_t}\\
  Y &\;=\; \mu \;+\; \beta_L\, L \;+\; \beta_T\, T \;+\; \beta_G\, G \;+\; \varepsilon,
       \qquad \varepsilon \sim \mathcal N\!\left(0,\ \sigma^{2}\right),
  \label{eq:fnd:dag_y}
\end{align}
%
with $L, G \sim \mathcal N(0,1)$. The planted causal effect is $\beta_T$, and it is exactly
\eur{\nbZeroTrueDag}: that is the grade every estimator in this chapter is marked against. The
baseline level $\mu$ plays no part in any of it --- it cancels from every difference --- which
is itself worth noticing: a causal contrast is invariant to the level of the outcome, and only
the \emph{coefficients} matter.

This world inherits the decision of \cref{sec:fnd:decision}: each email costs \eur{\nbZeroCost}
to send, so the only question that matters commercially is whether the effect clears that bar.
The planted \eur{\nbZeroTrueDag} does not. Whether an estimator \emph{sees} that it does not is
what \cref{sec:fnd:classical} puts to the test, and \cref{tab:fnd:strata} states the verdict in
euros.

\Cref{tab:fnd:params} gives every constant in
\crefrange{eq:fnd:uplift}{eq:fnd:dag_y}, its value, and --- more to the point --- why it has
that value rather than another. A constant baked into an equation with no justification is a
number the reader cannot argue with, and this book prefers numbers the reader can argue with.

\input{build/tables/nb00_params}

\subsection*{The zero-effect world, for the collider}

\Cref{sec:fnd:collider} needs a world where the true effect is \emph{exactly zero} and the
treatment is randomized. Then any effect an estimator reports is manufactured, not merely
mismeasured:
%
\begin{equation}
  T \sim \text{Bernoulli}(\tfrac12), \qquad
  Y = \mu_c + 0 \cdot T + \varepsilon, \qquad
  C = \mathbf 1\big\{\,\lambda_T\,T + \lambda_Y\,\tilde Y + \eta > c\,\big\},
  \label{eq:fnd:collider}
\end{equation}
%
Here $\tilde Y$ is spend standardized to unit variance and $\eta$ is standard normal noise
(constants in Appendix A). The coefficient on $T$ is \textbf{zero, by construction}: the email
does nothing. $C$ is ``responded to the campaign'' --- more likely if the customer was emailed
\emph{and} more likely if she was going to spend a lot. It is a common effect of treatment and
outcome: a collider. \Cref{eq:fnd:collider} is the smallest machine that manufactures a causal
effect from nothing, and \cref{sec:fnd:collider} prices it.

% =====================================================================================
\section{Identification}
\label{sec:fnd:ident}

\emph{Identification} is the question of whether the causal quantity could be recovered from the
data at all, given infinite data and a perfect estimator. It is prior to, and independent of,
estimation: no sampler, no machine-learning model and no quantity of data repairs an
identification failure, and \cref{sec:fnd:wrong} measures exactly that.

Without randomization the effect is not identified by the data alone. It is identified by the
data \emph{plus assumptions}, and those assumptions are the price of admission. Four are stated
here. Each is either tested later in this chapter or explicitly declared untestable --- an
assumption with no accompanying test is a wish.

\begin{assumption}[SUTVA: no interference, one version of treatment]\label{as:fnd:sutva}
\textbf{In plain words:} a customer's outcome depends on \emph{her own} email and on nothing
else --- not on whether her neighbour was emailed --- and ``the email'' means one single,
well-defined thing.

Two ways it fails, both concrete. \emph{Interference}: an emailed customer forwards the offer
to a friend who was \emph{not} emailed, and the friend buys. The friend's outcome now depends
on somebody else's treatment, so she no longer has just two potential outcomes and
\cref{eq:fnd:consistency} is not even well posed. \emph{Multiple versions}: ``emailed'' was in
fact three different creatives with three different offers, so $Y(1)$ is not a single number
but three, and the ``effect'' being estimated is an average over an undeclared mix.

Formally, unit $i$'s potential outcomes depend only on $T_i$ \citep{rubin1980, imbens2015}.
\textbf{Untestable from the data}, and named as such: it is discharged by knowing how the
campaign ran, not by a statistic. \Cref{ch:geo} plants a violation in a simulator, which is the
only place one can be seen.
\end{assumption}

\begin{assumption}[Unconfoundedness]\label{as:fnd:unconf}
\textbf{In plain words:} among customers who look the same on the measured features, who got
the email is as good as a coin flip.

That is, whatever drove the targeting is \emph{already in $X$}. Concretely: among customers
with the same recency, frequency and engagement, knowing that one of them happened to be
emailed tells you nothing further about what she would have spent. And the violation, equally
concretely: the marketer also had a hunch --- an unlogged ``about to churn'' signal --- that
drove both whom she emailed and what those customers were going to spend anyway. That signal is
not in $X$, so within-$X$ groups are still not comparable, and no estimator fitted on $X$ can
know it. Formally,
%
\begin{equation}
  \{\,Y(0),\, Y(1)\,\} \;\perp\; T \;\;\big|\;\; X .
  \label{eq:fnd:unconf}
\end{equation}
%
This is the load-bearing assumption of the entire chapter, and it is \textbf{untestable}: it is
a statement about potential outcomes that were never observed. \Cref{sec:fnd:wrong} does the
only honest thing available --- it plants a violation and measures what it costs
(\cref{fig:fnd:hidden}) --- and \Cref{ch:up} prices the strength a hidden confounder would need
to overturn a decision.
\end{assumption}

\begin{assumption}[Positivity, or overlap]\label{as:fnd:overlap}
\textbf{In plain words:} every kind of customer could have gone either way. Nobody is emailed
with certainty, and nobody is excluded with certainty.
%
\begin{equation}
  0 \;<\; e(x) \;<\; 1, \qquad e(x) \;=\; \Prob(T = 1 \mid X = x),
  \label{eq:fnd:overlap}
\end{equation}
%
where $e(x)$ is the \emph{propensity score} \citep{rosenbaum1983}: the probability that a
customer with features $x$ gets emailed. If some region of feature space is \emph{always}
emailed --- say every customer with a loyalty score above some level --- then in that region
there are no untreated counterparts, no like-for-like comparison exists, and no reweighting can
invent one. Unlike \cref{as:fnd:unconf}, this one is \textbf{testable}, because $e(x)$ is
estimable from observed data. It is tested in \cref{sec:fnd:valid}, where it passes
(\cref{fig:fnd:overlap}), and again in \cref{sec:fnd:real}, where it fails badly
(\cref{fig:fnd:lalonde}).
\end{assumption}

\begin{assumption}[A valid adjustment set]\label{as:fnd:backdoor}
\textbf{In plain words:} the variables we adjust for must close every back door, and must not
open a new one.

Using the vocabulary of \cref{sec:fnd:vocab}: $X$ blocks every backdoor path from $T$ to $Y$,
and contains \emph{no descendants of the treatment} --- no mediators, no colliders. This is
Pearl's \emph{backdoor criterion} \citep{pearl1995, pearl2009}. It is why
``adjust for everything in the warehouse'' is not a strategy: it closes the backdoor paths and
simultaneously opens the collider ones. It is \textbf{partially testable} --- the graph itself
is an assumption, but the \emph{consequence} of getting it wrong is measurable, and
\cref{sec:fnd:collider} measures it on a treatment whose true effect is exactly zero.
\end{assumption}

Under \crefrange{as:fnd:sutva}{as:fnd:backdoor}, the ATE is identified by the adjustment formula
\cref{eq:fnd:gformula} --- which Appendix~1.A defines, and which \cref{tab:fnd:strata} now
executes \emph{by hand}, once, before any model is fitted.

\subsection*{The g-formula, executed by hand}

Take the one-confounder world of \crefrange{eq:fnd:dag_t}{eq:fnd:dag_y}, where the planted
effect is \eur{\nbZeroTrueDag}. Cut loyalty $L$ into five quintiles --- five bins of
``customers who look alike'' --- and inside each bin compute the plain emailed-minus-not
difference. That is the inner bracket of \cref{eq:fnd:gformula}. Then average the five
differences, weighting each by how many customers it contains. That is the outer sum. The
result is \cref{tab:fnd:strata} and \cref{fig:fnd:strataplot}, and three numbers in them tell
the whole story:

\begin{itemize}[leftmargin=*]
  \item \textbf{Within each quintile}, the difference is about \eur{\nbZeroTrueDag} --- the
        planted truth. Like-for-like comparison recovers the effect, with no model, no
        estimator and no machinery.
  \item \textbf{The strata-weighted average} is \eur{\nbZeroStratAte}: the g-formula's answer,
        computed by hand.
  \item \textbf{The pooled contrast on the same data} is \eur{\nbZeroPooledNaive} --- far above
        both. And the \emph{\% emailed} column says exactly why: the emailed share climbs from
        \pc{\nbZeroStratLoEmailed} in the least loyal quintile to \pc{\nbZeroStratHiEmailed} in
        the most loyal one. The pooled number is comparing loyal emailed customers against
        casual un-emailed ones.
\end{itemize}

Read in euros, that gap is the whole decision. Against the \eur{\nbZeroCost} contact cost, the
pooled \eur{\nbZeroPooledNaive} clears the bar and says \emph{send to everyone}; the g-formula's
\eur{\nbZeroStratAte} sits below it and says \emph{send to no one}. The backdoor path is not a
statistical nicety --- it moves the number across the only threshold anyone acts on.

\input{build/tables/nb00_strata}
\input{build/floats/nb00_strata_plot}

The right-hand panel of \cref{fig:fnd:strataplot} is the mechanism, made visible: that climbing
emailed-share \emph{is} the confounding. Every rung of the ladder in \cref{sec:fnd:classical} is
a smarter, smoother way of doing precisely what \cref{tab:fnd:strata} does with five bins.

% =====================================================================================
\section{The graph: seeing, doing, and what to condition on}
\label{sec:fnd:collider}

Why draw graphs at all? Because \cref{as:fnd:unconf} and \cref{as:fnd:backdoor} both require
knowing \emph{which} variables belong in $X$ --- and ``adjust for the confounders'' is circular
until something says what a confounder is. The DAG is that something. It converts an abstract
assumption into an actionable rule, and it shows that the tempting instinct --- control for
everything we have --- \emph{backfires}. This section states the rule, executes it
(\cref{tab:fnd:strata} was the execution), and then prices the cost of getting it wrong.

\subsection*{Seeing is not doing}

\begin{intuition}{The barometer}
A falling barometer reliably precedes a storm. \emph{Seeing} a low reading is therefore
genuinely informative --- it tells you about the atmospheric pressure that drives both the
needle and the weather. Now go and \emph{force} the needle down with your thumb. Nothing
happens to the weather. The association was real; the intervention does nothing, because your
thumb severed the link to the common cause. That is the whole of confounding, in one image:
seeing and doing are different questions, and only one of them is what a decision needs.
\end{intuition}

\emph{Observing} $X = x$ selects the units that happened to arrive at $x$, carrying every reason
they got there. \emph{Doing} $do(X = x)$ reaches in and \emph{sets} it, overriding those reasons
\citep{pearl2009}. Every commercial decision is a $do$: ``should \emph{we} send the email?''
sets $T = 1$ by fiat. Every dashboard is a \emph{seeing}: ``what do emailed customers look
like?'' The two are routinely reported as if they were the same number, and the gap between them
is the most expensive error in the field.

Graphically, $do(T = t)$ is \textbf{surgery}: delete every arrow pointing \emph{into} $T$ --- the
natural causes of $T$ are overridden, because we, not they, are now choosing it --- then fix
$T = t$ and read off $Y$. Observing leaves the graph intact and merely filters.
\Cref{fig:fnd:surgery} draws both.

\input{build/floats/nb00_do_surgery}

This makes the value of randomization vivid: \textbf{a coin flip performs the surgery
physically}. It severs $L \rightarrow T$ in the real world, not on paper, which is why an
experiment needs no adjustment and no graph. Observational adjustment is the attempt to
reconstruct \emph{doing} from \emph{seeing}, by conditioning on the arrows the coin flip would
have cut.

And in the one-confounder world we can put a number on the difference, because we know the
graph exactly. There is precisely \emph{one} backdoor path, $T \leftarrow L \rightarrow Y$:
loyalty raises the send probability (through $\alpha_L$) and raises spend (through $\beta_L$).
The trait $G$ is not on any backdoor path --- it moves only $Y$ --- and is a deliberate foil.
So:
%
\[
  \underbrace{\eur{\nbZeroPooledNaive}}_{\text{seeing}}
  \;=\; \underbrace{\eur{\nbZeroTrueDag}}_{\text{doing } (\beta_T)}
  \;+\; \underbrace{\eur{\nbZeroBackdoorGap}}_{\text{the backdoor, through } L} .
\]
%
The gap \emph{is} the backdoor path --- not a bias term, not an error, but the association
flowing along $T \leftarrow L \rightarrow Y$, quantified. Blocking that path is what
\cref{tab:fnd:strata} did with its five quintiles, and it closed the \eur{\nbZeroBackdoorGap}
back down to the planted \eur{\nbZeroTrueDag}. \Cref{fig:fnd:seeingdoing} is the same statement
as a picture. When the backdoor variables are measured, the backdoor adjustment formula
converts one into the other,
%
\begin{equation}
  \Prob\big(y \mid do(T{=}t)\big) \;=\; \sum_x \Prob\big(y \mid T{=}t,\, X{=}x\big)\, \Prob(X{=}x),
  \label{eq:fnd:backdoor}
\end{equation}
%
which is \cref{eq:fnd:gformula} written for distributions rather than means.

\input{build/floats/nb00_seeing_doing}

\subsection*{Three graphs, and only one calls for adjustment}

Appendix~1.A defines the three structures; here is what conditioning does to each, with the
algebra, on the email example (\cref{fig:fnd:dags} drew them).

\begin{description}[leftmargin=*, style=nextline]
  \item[Confounder --- a backdoor path: $T \leftarrow L \rightarrow Y$. \textmd{\textbf{Adjust.}}]
    Loyalty is a common cause. Regress $Y$ on $T$ alone in \cref{eq:fnd:dag_y} and the
    coefficient comes out at \eur{\nbZeroPooledNaive}, because $T$ is standing in for the $L$ it
    is correlated with. Put $L$ in the regression and the coefficient is $\beta_T =
    \eur{\nbZeroTrueDag}$. Conditioning is licensed because it delivers exactly what
    \cref{as:fnd:unconf} asks for: $\{Y(0), Y(1)\} \perp T \mid L$.
  \item[Mediator --- on the causal path: $T \rightarrow M \rightarrow Y$. \textmd{\textbf{Do not adjust.}}]
    Let $M$ be ``opened the email''. The email works \emph{through} being opened, so the path
    $T \rightarrow M \rightarrow Y$ carries a genuine part of the effect. Conditioning on $M$
    blocks it and removes that part from the estimate: the \emph{total} effect is no longer
    identified. Note that ``opened'' is a column in every marketing warehouse, which is what
    makes this the most tempting error in the list. \Cref{ch:med} shows how to split an effect
    into direct and indirect parts when that split is genuinely what is wanted.
  \item[Collider --- a common effect: $T \rightarrow C \leftarrow Y$. \textmd{\textbf{Do not condition.}}]
    Let $C$ be ``responded''. The path $T \rightarrow C \leftarrow Y$ is \emph{blocked by
    default} --- $T$ and $Y$ are not associated through it. Conditioning on $C$ \emph{opens} it
    and manufactures association out of nothing. This is not a subtle inefficiency; it is the
    subject of the next subsection, and it costs real money.
\end{description}

\subsection*{The collider, priced}

The most common self-inflicted wound in marketing analytics is analysing only the customers who
responded --- opened, clicked, replied, converted. \Cref{eq:fnd:collider} makes the damage
visible by planting an email whose true effect is \textbf{exactly zero} and \emph{randomizing}
it, so that even the naive comparison is unbiased and nothing else can be blamed. On all
\nbZeroColliderN{} customers, the estimated email coefficient is \eur{\nbZeroColliderAll} ---
correctly, nothing. Restricted to the \pc{\nbZeroColliderShare} who responded, it is
\eur{\nbZeroColliderResp}. \Cref{fig:fnd:collideraway} shows both: no gap in spend by email
among all customers, and a spurious negative gap once the sample is cut to responders only.

An effect out of thin air, and with the wrong sign.

\begin{intuition}{Explaining away}
Among responders only, \emph{having been emailed already explains the response}. So an emailed
responder needed \emph{less} spending appetite to cross the response threshold than an
un-emailed responder did --- the un-emailed one had to get there on appetite alone.
Conditioning on the common effect therefore induces a \emph{negative} association between its
two causes, which \citet{pearl2009} calls \emph{explaining away}: learning that one cause is
present makes the other less necessary. The email did nothing; the selection into ``responder''
did all of it.
\end{intuition}

\input{build/floats/nb00_collider_explaining_away}

Scaled across the \nbZeroColliderNResp{} responders, the responders-only analysis reports
\eur{\nbZeroColliderPhantom} of phantom revenue destruction on a campaign whose true effect is
zero. And nothing about that analysis \emph{looks} broken: the treatment was randomized, the
sample is large, the coefficient is precisely estimated, the standard error is small. It is
simply an answer to a question nobody asked.

The rule this yields is short and absolute: \textbf{post-treatment variables stay out of the
adjustment set}. Opens, clicks, conversions and responses are all descendants of the treatment,
and \cref{as:fnd:backdoor} excludes them.

\subsection*{M-bias: the collider you cannot see}

The uncomfortable corollary --- and the reason \cref{as:fnd:backdoor} is only \emph{partially}
testable --- is that a collider need not be post-treatment, and need not even be visible in the
drawn graph. Consider the classic M-shaped structure \citep{elwert2014, ding2015}, with two
unobserved variables $U_1$ and $U_2$ and one observed \emph{pre-treatment} covariate $Z$:
%
\begin{equation}
  U_1 \rightarrow Z \leftarrow U_2, \qquad U_1 \rightarrow T, \qquad U_2 \rightarrow Y .
  \label{eq:fnd:mbias}
\end{equation}
%
Here $Z$ is measured \emph{before} treatment, is correlated with both $T$ and $Y$, and passes
every screening rule a practitioner is likely to apply. And yet the path
$T \leftarrow U_1 \rightarrow Z \leftarrow U_2 \rightarrow Y$ is \emph{blocked} --- $Z$ is a
collider on it --- so the naive comparison is unconfounded, and conditioning on $Z$
\emph{opens} the path and \emph{creates} bias. Adjusting for a pre-treatment covariate can make
things worse.

Two things follow. First, ``adjust for everything measured before treatment'' is not a safe
rule; it is merely safer than adjusting for everything. Second, the graph is a claim about the
world, not a picture of the data, and no test on the observed variables distinguishes
\cref{eq:fnd:mbias} from a world in which $Z$ is an honest confounder. The punchline is plain: a
pre-treatment variable that passes every screen can still be a collider. \Cref{ch:ctrl} devotes a
chapter to the problem and demonstrates M-bias with a working example.

% =====================================================================================
\section{The classical baseline: the estimator ladder}
\label{sec:fnd:classical}

The discipline of this book is to ask, before any sampler, what a competent analyst produces
\emph{without} one: a point estimate, and an interval whose covariance assumption is stated out
loud. This section is that read, in full, on the one-confounder world of
\crefrange{eq:fnd:dag_t}{eq:fnd:dag_y}. There are no priors here, no likelihood and no chains.
\Cref{ch:pvp} adds them --- and its first finding is that they change nothing.

Let $\hat m(t, x) \approx \E[Y \mid T{=}t, X{=}x]$ be a fitted outcome model,
$\hat e(x) \approx \Prob(T{=}1 \mid X{=}x)$ a fitted propensity, and $\bar Y_1, \bar Y_0$ the raw
arm means over $n$ customers. The four rungs are
%
\begin{align}
  \hat\tau_{\text{naive}} &\;=\; \bar Y_1 - \bar Y_0, \label{eq:fnd:rung_naive}\\[2pt]
  \hat\tau_{\text{reg}}   &\;=\; \frac1n \sum_{i=1}^{n}
     \big[\hat m(1, x_i) - \hat m(0, x_i)\big], \label{eq:fnd:reg}\\[2pt]
  \hat\tau_{\text{IPW}}   &\;=\; \frac1n \sum_{i=1}^{n}
     \left[\frac{T_i\,Y_i}{\hat e(x_i)} - \frac{(1 - T_i)\,Y_i}{1 - \hat e(x_i)}\right],
     \label{eq:fnd:ipw}\\[2pt]
  \hat\tau_{\text{AIPW}}  &\;=\; \frac1n \sum_{i=1}^{n}
     \left[\hat m(1, x_i) - \hat m(0, x_i)
      + \frac{T_i\big(Y_i - \hat m(1, x_i)\big)}{\hat e(x_i)}
      - \frac{(1 - T_i)\big(Y_i - \hat m(0, x_i)\big)}{1 - \hat e(x_i)}\right].
     \label{eq:fnd:aipw}
\end{align}

Each is a way of executing \cref{eq:fnd:gformula}, and each buys its correctness with a
different assumption. Taking them in turn:

\Cref{eq:fnd:rung_naive} is \cref{eq:fnd:naive}: the dashboard. It ignores $X$ entirely, and it
is on the ladder as the control condition --- the rung that fails.

\Cref{eq:fnd:reg} is the g-formula with the five quintiles of \cref{tab:fnd:strata} replaced by
a \emph{fitted line}: instead of averaging within bins, predict each customer's outcome under
both treatments and average the difference over the whole file. Note it predicts for
\emph{every} customer under \emph{both} arms --- including the arm she was not in --- which is
precisely the imputation of the missing half of \cref{tab:fnd:oracle}. It is correct if the
outcome model is correct.

\Cref{eq:fnd:ipw} attacks from the other side. Weight each treated customer by
$1/\hat e(x_i)$: a customer of a kind that is \emph{rarely} emailed has a small $\hat e$, so
she counts for a lot, standing in for all the customers like her who were not emailed.
Concretely, a rarely-emailed customer with estimated propensity $\hat e = \nbZeroIpwEhat$ carries
a weight of $1/\nbZeroIpwEhat \approx \nbZeroIpwWeight$, so she stands in for about
\nbZeroIpwWeight{} customers of her kind. The reweighted treated arm is thereby made to look like
the \emph{population}'s feature mix rather
than the treated arm's \citep{rosenbaum1983} --- which is exactly the substitution
\cref{eq:fnd:gformula} demands and \cref{eq:fnd:naive} gets wrong. It is correct if the
propensity model is correct.

\begin{intuition}{Why ``doubly'' robust}
\Cref{eq:fnd:aipw} is the regression answer \emph{plus} a weighted-residual correction, and the
correction is the clever part. If $\hat m$ is right, the residuals $Y - \hat m$ are mean-zero
noise, so the correction term vanishes in expectation and you keep \cref{eq:fnd:reg}. If
instead $\hat m$ is \emph{wrong} but $\hat e$ is right, that same correction is an unbiased
inverse-propensity estimate of precisely the error $\hat m$ is making --- and it repairs it.
\textbf{Either model being correct suffices}, and you do not have to know which one it is
\citep{robins1994}. Two shots at the target, and you only need one to land.
\end{intuition}

The implementation makes three design choices, each a choice rather than a default:

\begin{itemize}[leftmargin=*, itemsep=.2em]
  \item \textbf{Cross-fit} \citep{chernozhukov2018} over \nbZeroAipwNFolds{} folds. Fit $\hat m$
        and $\hat e$ on all but one fold and score the held-back fold, rotating until every
        customer is scored by a model that never saw her. This removes the overfitting bias an
        adaptive learner would inject into the correction term.
  \item \textbf{Winsorise the propensity} at $[\nbZeroAipwTrim,\, 1 - \nbZeroAipwTrim]$. This
        bounds the weights, it does not filter the sample; no unit is dropped, so the estimand
        stays the full-population ATE.
  \item \textbf{Use a flexible outcome model}: a gradient-boosted tree ensemble, not
        \cref{eq:fnd:reg}'s straight line. When \cref{eq:fnd:aipw} and \cref{eq:fnd:reg} then
        agree, a flexible learner and a linear one have agreed --- evidence about the
        specification, not just the arithmetic.
\end{itemize}

\subsection*{The standard error each rung demands}

The standard-error column of \cref{tab:fnd:ladder} is not bookkeeping: an interval built on the
wrong covariance is a confident lie. The four rungs demand four different answers --- a Welch
error for the naive difference, an HC1 heteroskedasticity-robust error for the regression, and a
refit bootstrap for IPW and AIPW, whose propensity is itself estimated. Appendix~1.A justifies
each choice and measures what the robust default costs when the errors are in fact well behaved
(it is negligible). The result table is next.

\input{build/tables/nb00_ladder}

\Cref{tab:fnd:ladder} is the chapter's central result. On one confounded sample of
\nbZeroNCustomers{} customers with a planted effect of \eur{\nbZeroTrueDag}, the naive rung
returns \eur{\nbZeroLadNaive} --- \eur{\nbZeroLadNaiveErr} off the planted effect --- and its
90\,\% interval, $[\eur{\nbZeroLadNaiveLo},\ \eur{\nbZeroLadNaiveHi}]$, does not come close to
containing the truth. Regression adjustment returns \eur{\nbZeroLadReg}, inverse-propensity
weighting \eur{\nbZeroLadIpw}, and the doubly-robust estimator \eur{\nbZeroLadAipw}. Every
adjusted rung lands within \eur{\nbZeroLadAdjWorst} of the planted effect. Same data, four
estimators, and the one that ignores the graph is the only one that fails: \textbf{the estimator
is the identification strategy}, and the arithmetic is a detail by comparison.

One feature of \cref{tab:fnd:ladder} is the most misread property of an interval. On this seed,
two of the three adjusted intervals exclude the planted \eur{\nbZeroTrueDag}. The HC1 interval,
$[\eur{\nbZeroLadRegLo},\ \eur{\nbZeroLadRegHi}]$, and the AIPW interval,
$[\eur{\nbZeroLadAipwLo},\ \eur{\nbZeroLadAipwHi}]$, both sit below it; IPW's
$[\eur{\nbZeroLadIpwLo},\ \eur{\nbZeroLadIpwHi}]$ covers. This is not a defect, and it is not two
independent one-in-ten events. The three rungs read the \emph{same} \nbZeroNCustomers{}
customers. A sample that sits low sits low for all of them at once, so they miss \emph{together},
and \Cref{ch:pvp}'s credible interval misses on the same panel for the same reason. One unlucky
draw seen three times is one unlucky draw. A 90\,\% interval is supposed to miss about one time
in ten; intervals that never missed would be too wide to use. So do not distrust the estimator on
one panel. Audit the \emph{procedure} over many fresh ones --- \nbZeroNSeeds{} of them in
\cref{tab:fnd:multiseed}, the next section.

% =====================================================================================
\section{Diagnostics and validation}
\label{sec:fnd:valid}

Three diagnostics are run here, and each is taught the same way: what it is, what it computes,
what it would look like if it \emph{failed}, and only then what it says on this data. A
diagnostic reported without its failure mode is a decoration.

\subsection*{Bias is what survives repetition}

\textbf{What it is.} An estimator is not graded on one draw of the world. \emph{Unbiased} is a
statement about the \emph{average over repeated samples}, so the honest test regenerates the
data from scratch, refits everything, and asks where the estimates centre.

\textbf{What it computes.} Draw \nbZeroNSeeds{} fresh panels from
\crefrange{eq:fnd:dag_t}{eq:fnd:dag_y}, refit all four rungs on each, and record the mean of each
rung's estimates. Bias is that mean minus the planted \eur{\nbZeroTrueDag}.

\textbf{The yardstick, without which the number is meaningless.} An average over finitely many
panels carries its own noise. So the table reports, beside each mean, its \emph{Monte-Carlo
standard error} --- the standard deviation of the estimates divided by $\sqrt{\nbZeroNSeeds}$ ---
and the ratio of the measured bias to it. That ratio is the column to read.

\begin{intuition}{Why the Monte-Carlo error is the whole test}
An unbiased estimator run on \nbZeroNSeeds{} panels will \emph{not} average to exactly the
truth; it will land a Monte-Carlo standard error or two away, because \nbZeroNSeeds{} is not
infinity. So ``the mean is not exactly \eur{\nbZeroTrueDag}'' proves nothing on its own. The
question is whether the miss is \emph{bigger than the noise in the measurement of the miss}. A
bias of two Monte-Carlo standard errors is what an unbiased estimator looks like. A bias of
eighty is a broken identification argument. Without the yardstick printed beside it, both are
just small numbers on a page.
\end{intuition}

\textbf{What it says.} The two failure modes separate with no ambiguity at all:

\begin{itemize}[leftmargin=*]
  \item The naive rung averages \eur{\nbZeroMsNaiveMean} across panels --- a bias of
        \eur{\nbZeroMsNaiveBias} against a Monte-Carlo error of \eur{\nbZeroMsNaiveMcse}, which
        places it \nbZeroMsNaiveBiasInMcse{} Monte-Carlo standard errors from zero. Nothing about
        that is an artefact of having drawn \nbZeroNSeeds{} panels rather than more. It is the
        confounding, and confounding is a property of the design, not of $n$.
  \item Every adjusted rung centres on the truth: regression at \eur{\nbZeroMsRegMean}, IPW at
        \eur{\nbZeroMsIpwMean}, AIPW at \eur{\nbZeroMsAipwMean}, against the planted
        \eur{\nbZeroTrueDag} --- each within a couple of Monte-Carlo standard errors of it
        (\nbZeroMsRegBiasInMcse, \nbZeroMsIpwBiasInMcse{} and \nbZeroMsAipwBiasInMcse{}
        respectively), which is exactly what a finite panel count does to an unbiased estimator.
        Their sample-to-sample standard deviations (\eur{\nbZeroMsRegSd}, \eur{\nbZeroMsIpwSd},
        \eur{\nbZeroMsAipwSd}) price what each costs in \emph{precision}, and IPW --- which
        throws away information by not modelling the outcome at all --- is measurably the
        noisiest of the three.
  \item The AIPW 90\,\% confidence interval covered the planted effect on \nbZeroAipwCovered{}
        of \nbZeroNSeeds{} panels: \pc{\nbZeroAipwCovPct}, against a nominal 90. The procedure
        keeps its promise. This is the \emph{only} audit a confidence interval admits, and
        \Cref{ch:pvp} explains why.
\end{itemize}

\input{build/tables/nb00_multiseed}
\input{build/floats/nb00_rungs}

\subsection*{Overlap: the assumption that can be tested}

\textbf{What it is.} \Cref{as:fnd:overlap} says every kind of customer could have gone either
way. Unlike unconfoundedness, this is a statement about \emph{observable} quantities, so it can
be checked --- and it is the only one of the four that can.

\textbf{What it computes.} Fit the propensity $\hat e(x)$, then look at its range in each arm. If
some region of feature space is occupied by only one arm, positivity fails there.

\textbf{A subtlety that decides whether the test can fail at all.} The IPW rung bounds $\hat e$
into $[0.05, 0.95]$ to keep its weights finite. A diagnostic computed on that \emph{clipped}
vector would report the clip back to itself --- its extremes would be the clip bounds by
construction --- and could never fail at precisely the boundary it is supposed to police.
\Cref{fig:fnd:overlap} therefore plots the \textbf{raw} score. A test that cannot fail is not a
test.

\textbf{What it says.} The emailed arm spans $[\nbZeroETrtMin, \nbZeroETrtMax]$ and the control
arm $[\nbZeroECtlMin, \nbZeroECtlMax]$ --- the same region, occupied by both --- and across the
whole sample the clip would bind on only \nbZeroClipN{} of \nbZeroNCustomers{} customers
(\pc{\nbZeroClipFrac}). Overlap holds, and the verdict was reached by a test that was allowed to
fail. \Cref{sec:fnd:real} shows the same test failing loudly on real data.

\subsection*{Balance: did the weights actually do their job?}

\textbf{What it is.} IPW claims to make the treated arm resemble the population. Balance is the
check on that claim. Its statistic is the \emph{standardized mean difference} (SMD): the gap in a
feature's mean between the arms, measured in pooled standard deviations, with a conventional
balance bar of $|\text{SMD}| < 0.1$ (\cref{eq:fnd:smd}, Appendix~1.A).

\textbf{What it says.} In the raw data, loyalty's SMD between the arms is \nbZeroSmdRaw{} ---
far above the $0.1$ bar, which is confounding, visible. Under the inverse-propensity weights it
collapses to \nbZeroSmdIpw. That collapse \emph{is} what an IPW estimator does when it works,
and it is the diagnostic to run before believing one.

\input{build/floats/nb00_overlap}

\begin{remark}
Note the asymmetry the three diagnostics reveal, because it is the shape of the whole field.
Positivity and balance can be checked, and were. Unconfoundedness (\cref{as:fnd:unconf}) cannot
be checked at all, and is the assumption everything actually rests on. \textbf{A causal analysis
that reports only the testable diagnostics has reported only the assumptions it was not worried
about} --- and \cref{sec:fnd:wrong} shows what that looks like when the untestable one is false.
\end{remark}

% =====================================================================================
\section{What can go wrong}
\label{sec:fnd:wrong}

Everything so far has been a success story, and that is a dangerous thing to teach: the ladder
recovered \eur{\nbZeroTrueDag} because the one confounder was \emph{in the data}. The
foundational question is what happens when it is not.

\subsection*{The unmeasured backdoor}

\Cref{fig:fnd:hidden} and \cref{tab:fnd:hidden} rerun the entire ladder on data in which a hidden
trait $U$ --- a salesperson's hunch, an unlogged loyalty signal --- is wired into both the
assignment and the outcome, while every rung is fitted on the five \emph{observed} features only,
exactly as a real analyst would have to. The naive rung returns \eur{\nbZeroHidNaive} against a
truth of \eur{\nbZeroHiddenTrue}. Regression returns \eur{\nbZeroHidReg}, IPW
\eur{\nbZeroHidIpw}, and AIPW \eur{\nbZeroHidAipw}.

Read that row again. Every adjusted rung sits at least \eur{\nbZeroHiddenMinBias} above the
truth, and \emph{they fail together}. This is the most important negative result in the chapter,
and it has two parts.

\textbf{First, double robustness offers no protection here}, and the reason is precise rather
than empirical. AIPW insures against a mis-specified \emph{model of the observed features}. It
has no insurance to offer against a feature that is \emph{absent}: both of its two models are
fitted on the same $X$, and the open backdoor runs through $U$, which neither can see. Two
chances at the wrong problem are not one chance at the right one.

\textbf{Second, the failure is silent} --- and that is a measurement, not a rhetorical flourish.
Run the complete diagnostic sheet of \cref{sec:fnd:valid} on one of these hidden-$U$ panels, and
\emph{every line passes}:

\begin{itemize}[leftmargin=*]
  \item \textbf{Overlap.} The raw propensity spans $[\nbZeroHidEMin, \nbZeroHidEMax]$, with
        \nbZeroHidClipN{} customers outside the $[0.05, 0.95]$ clip. Positivity holds.
  \item \textbf{Balance.} The worst SMD among the five measured features falls from
        \nbZeroHidSmdRaw{} to \nbZeroHidSmdIpw{} under the weights --- comfortably inside the
        conventional $0.1$ bar. The weights are doing their job.
  \item \textbf{Precision.} The doubly-robust estimate is \eur{\nbZeroHidPanelEst}, with a
        standard error of \eur{\nbZeroHidPanelSe} and a 90\,\% interval of
        $[\eur{\nbZeroHidPanelLo},\ \eur{\nbZeroHidPanelHi}]$. Tight, and converged.
\end{itemize}
%
Every box is ticked. The planted truth on that panel is \eur{\nbZeroHidPanelTrue}, so the
estimate is \eur{\nbZeroHidPanelBias} out. \textbf{A confidently wrong answer is the default
output of a well-run analysis on a badly-drawn graph}, and \Cref{ch:pvp} shows that a Bayesian
posterior fails in exactly the same silent way, with the same serenity. The reason no diagnostic
caught it is structural, and it is worth stating as a rule: every test in \cref{sec:fnd:valid}
interrogates the features that \emph{are} in the data, and the assumption that failed
(\cref{as:fnd:unconf}) is a statement about one that is not.

\input{build/floats/nb00_hidden_u}
\input{build/tables/nb00_hidden}

And it is not an abstraction, because these panels come from the same simulator, the same
customer base and the same \eur{\nbZeroCost} contact cost as \cref{sec:fnd:euro}. The planted
effect is \eur{\nbZeroHiddenTrue}, which is \emph{below} the cost: the correct decision is to
send nothing. Every adjusted rung reports an effect above the cost --- AIPW at
\eur{\nbZeroHidAipw}, regression at \eur{\nbZeroHidReg}, IPW at \eur{\nbZeroHidIpw} --- and
therefore recommends \nbZeroHiddenVerdict: email everyone. Executed on the base, that
recommendation burns \eur{\nbZeroHiddenPolBurn}. It is the identical euro error as the confounded
dashboard of \cref{sec:fnd:decision}, arrived at by a doubly-robust estimator with a clean
diagnostic sheet and a tight interval. The badge is precisely what makes it dangerous: it is what
gets the number believed in the meeting.

The honest responses are three, and none of them is a better estimator: get better data; get a
better \emph{design} (randomize --- \cref{eq:fnd:rct}); or run a \emph{sensitivity analysis},
asking how strong a hidden $U$ would have to be to overturn the decision. \Cref{ch:up} prices
that stress as a tipping point and an E-value.

\subsection*{When positivity fails}

The second assumption fails differently, and worse. Keep the confounder fully observed, but
sharpen the selection until it is nearly deterministic --- so that high-loyalty customers are
almost always emailed and low-loyalty ones almost never. The propensity clip now binds on
\pc{\nbZeroPosClipBad} of customers, against \pc{\nbZeroPosClipOk} under the gentle selection of
\cref{eq:fnd:dag_t}, and IPW drifts from \eur{\nbZeroPosIpwOk} to \eur{\nbZeroPosIpwBad} against
a truth of \eur{\nbZeroTrueDag}. (Both shares are measured on the raw propensity, on the same
panels. The clipped share under gentle selection is small but not zero, and the difference
between ``small'' and ``zero'' is exactly the sort of claim this chapter refuses to make without
looking.)

The estimator is not being asked to model something badly. It is being asked to compare customers
who \emph{have no counterparts in the other arm}. No reweighting recovers a full-population ATE
from data that contains no such comparison, and the correct response is to change the estimand
--- to report an effect on the region where overlap holds --- rather than to extrapolate over the
region where it does not. \Cref{fig:fnd:overlap} is the diagnostic that catches this before the
number is trusted, and \cref{sec:fnd:real} is what it looks like when nobody ran it.

% =====================================================================================
\section{On real data: LaLonde}
\label{sec:fnd:real}

Everything above was graded against a truth that was planted. The obvious objection --- that
recovering an effect somebody put there proves nothing --- deserves a real answer, and causal
inference has a canonical one. This section is the second tier of \cref{sec:fnd:dgm}'s three:
real data, with a randomized benchmark, so an observational method can be \emph{graded} rather
than merely trusted. It is worked step by step, because it is the payoff of the whole chapter.

\subsection*{Step 1. The problem, and why this dataset can settle it}

\citet{lalonde1986} studied a 1970s job-training programme, the National Supported Work (NSW)
demonstration. Its decisive feature is that it \emph{included a randomized arm}, so the true
effect of training on 1978 earnings is known experimentally. On the subsample this book uses ---
the extract of \citet{dehejia1999} --- that experimental benchmark is about
\usd{\nbZeroLlBenchmark}; LaLonde's own headline on the full sample was a substantially smaller
figure, which is one reason the subsample matters. LaLonde then asked the question that founded a
literature. \textbf{Throw the experimental controls away.} Replace them with a
\emph{non-experimental} comparison group --- here the controls drawn from the Panel Study of
Income Dynamics (PSID), the only kind of comparison group an analyst without an experiment ever
has --- and see whether adjustment can recover the experimental answer.

This is grading on real data, and it is legitimate precisely because the benchmark came from
randomization. Note what is still missing: no individual's counterfactual is known here either.
What is known is the \emph{average}, and that is enough to score an estimator.

\subsection*{Step 2. Look at the data before estimating anything}

\nbZeroLlTreated{} trainees against \nbZeroLlControl{} PSID controls; treatment is job
training; the outcome is 1978 earnings; there are \nbZeroLlNCovs{} covariates (age, education,
race, marital status, and earnings in 1974 and 1975). \Cref{tab:fnd:llbalance} and
\cref{fig:fnd:llcov} show what \emph{selection} looks like when it is not simulated, and the
1974 and 1975 earnings rows carry the fact that decides the sign of everything downstream: the
trainees were already earning far less than the PSID controls, years \emph{before} any
training took place. That is $\E[Y(0) \mid T{=}1] \ll \E[Y(0) \mid T{=}0]$, measured directly.

The imbalance is not confined to earnings, and the table is worth reading rather than
summarising: the \emph{largest} standardized mean difference --- \cref{eq:fnd:smd}, the same
statistic as \cref{sec:fnd:valid} --- is \nbZeroLlSmdWorst{} on \nbZeroLlSmdWorstName{}, against
a conventional bar of $0.1$. These two groups are not comparable on any dimension, which is
precisely why the naive contrast between them is meaningless and the adjusted one is difficult.

\input{build/tables/nb00_ll_balance}
\input{build/floats/nb00_ll_covariates}

\subsection*{Step 3. Seeing: the naive difference says training \emph{destroyed} earnings}

The naive difference of means (\cref{eq:fnd:naive}) is \usd{\nbZeroLlNaive}. Not small.
\emph{Negative}. Read literally, job training cost these workers money.

This is \cref{eq:fnd:decomp} in the wild, and the decomposition explains the sign exactly. The
trainees' own baseline, $\E[Y(0) \mid T{=}1]$, is far \emph{below} the PSID controls'
$\E[Y(0) \mid T{=}0]$ --- \cref{tab:fnd:llbalance} shows it directly, in the 1974 and 1975
earnings rows. So the selection-bias term is large and \emph{negative}: these workers had a
head start in reverse. It swamps a positive training effect and reverses its sign. The very
same identity that inflated the email's effect by \eur{\nbZeroSelBias} in \cref{sec:fnd:vocab}
here \emph{hides} a real effect entirely.

\subsection*{Step 4. Doing: adjustment moves the estimate across zero}

Now adjust for the \nbZeroLlNCovs{} covariates --- \cref{eq:fnd:gformula}, executed by
regression, exactly as in \cref{eq:fnd:reg}, and licensed by \cref{as:fnd:backdoor} on the claim
that those covariates block the backdoor paths from training to earnings. The estimate moves by
\usd{\nbZeroLlSwing}, \textbf{across zero}, to \usd{\nbZeroLlAdj} --- right sign, right order of
magnitude, and within \usd{\nbZeroLlShortfall} of the randomized benchmark
(\cref{tab:fnd:lalonde}, \cref{fig:fnd:llbars}).

This is the whole chapter, on data nobody simulated: \textbf{the naive comparison is not merely
imprecise, it points the wrong way, and adjustment repairs the sign.}

\input{build/tables/nb00_lalonde}
\input{build/floats/nb00_ll_bars}

\subsection*{Step 5. Why it is not exact: positivity fails}

The remaining \usd{\nbZeroLlShortfall} is not a rounding error, and it is not left unexplained.
Run \cref{as:fnd:overlap}'s diagnostic --- the same one that passed in \cref{fig:fnd:overlap} ---
on LaLonde, and it fails loudly: \pc{\nbZeroLlCtrlLowE} of the PSID controls sit below
$e(x) = 0.10$, against \pc{\nbZeroLlTrtLowE} of the trainees (\cref{fig:fnd:lalonde}). Most of the
comparison pool looks nothing like a trainee.

A regression adjusting for covariates across that pool is not comparing like with like. It is
\emph{extrapolating} a fitted surface into a region where one arm is essentially absent, and
\cref{eq:fnd:gformula} was never licensed there --- the inner bracket
$\E[Y \mid T{=}1, X{=}x]$ is being evaluated at values of $x$ where almost no trainee exists.

\input{build/floats/nb00_lalonde_overlap}

\subsection*{Step 6. The repair, and what it does not buy}

That is why the field moved to propensity trimming and matching \citep{dehejia1999}: restrict the
comparison group to controls who could plausibly have been trainees, and the extrapolation goes
away. Whether the \emph{shortfall} goes away with it is a separate question, and this book is
obliged to answer it by running the estimator rather than by citing the paper.

Trimming to $0.05 < \hat e(x) < 0.95$ drops \nbZeroLlTrimDropped{} of the \nbZeroLlN{} rows,
leaving \nbZeroLlTrimN, and it does exactly what it advertises to the covariates: the worst
standardized mean difference falls from \nbZeroLlSmdFull{} to \nbZeroLlSmdTrim. The estimate on
the trimmed sample is \usd{\nbZeroLlAdjTrim} --- against \usd{\nbZeroLlAdj} untrimmed and a
benchmark of \usd{\nbZeroLlBenchmark}. \textbf{It moves sideways, not up.}

That result is more instructive than the tidy one would have been. A trim deletes the region in
which the regression was extrapolating; it does not \emph{reconstruct} the counterfactual that
was never sampled, and no amount of balance conjures a comparison group out of a survey that
contains nobody like a trainee. \Citet{dehejia1999} close the gap with a considerably richer
propensity specification and matching on their subsample --- a modelling argument, defended on
its merits, not a mechanical filter. So the honest lesson is narrower than the usual telling, and
more useful: \textbf{adjustment did most of the job; its own diagnostic correctly flagged the part
it could not do; and the diagnostic told the analyst \emph{where} the estimate was unsupported. It
did not, and cannot, say what the supported estimate would have been.}

\begin{remark}[Why this dataset, and not a marketing one]
A marketing analyst may reasonably ask what a job-training programme has to do with an email
campaign. The answer is that LaLonde is the only kind of dataset that can settle the argument: it
carries both a randomized benchmark and an observational comparison group, on the same treatment,
in the same population. That is the exam a causal method has to sit, and real marketing data ---
which has the observational arm and never the benchmark --- cannot set it. \Cref{ch:geo} sits the
same exam on a real, randomized geo experiment, and the shape of the finding is identical.
\end{remark}

% =====================================================================================
\section{The decision, in euros}
\label{sec:fnd:euro}

Statistical bias becomes a business problem the moment somebody acts on it. The simulator knows
every customer's true effect $\tau_i$, so each policy can be \emph{priced} on the actual customer
base --- and it is worth being precise about which half of that sentence uses the oracle. The
pricing does. The \emph{deciding} must not, or the exercise proves nothing about a method an
analyst could actually run.

\input{build/tables/nb00_policy}

\Cref{tab:fnd:policy} prices four policies on the same \nbZeroNCustomers{} customers, and the
rows are ordered by how much they know.

The confounded read-out --- \eur{\nbZeroNaiveObs} per contact against an \eur{\nbZeroCost} cost
--- says \emph{email everyone}. Doing so on this base realises \eur{\nbZeroPolEveryone}: it burns
\eur{\nbZeroPolBurn}.

The second row is the one an analyst can actually reach. Running the doubly-robust rung of
\cref{eq:fnd:aipw} on \emph{this} base --- the same five observed features, no oracle, no $\tau$
--- returns \eur{\nbZeroPolAdjEst}, with a 90\,\% interval of
$[\eur{\nbZeroPolAdjLo},\ \eur{\nbZeroPolAdjHi}]$, and lands it on the correct side of the cost.
It says \emph{email no one}, and it is right. The third row confirms that the true ATE of
\eur{\nbZeroAte} agrees, which is reassurance rather than evidence, since nobody in a real
company gets to look at it.

That ordering is the point of the table. The \eur{\nbZeroPolBurn} is not saved by knowing the
truth; it is saved by \emph{adjustment}, and the second row is the receipt. \textbf{The
identification error is not a footnote; it is a profit-and-loss line}, and its size here is
roughly \eur{\nbZeroPolBurn} per \nbZeroNCustomers{} contacts. The verdict is also less fragile
than it looks: the blanket send breaks even only at a contact cost below
\eur{\nbZeroBreakevenCost}, so at \eur{\nbZeroCost} it is not a close call --- the confounded
number does not merely exaggerate the effect, it moves it across the only threshold anyone cares
about.

The last row points at the rest of the book. An oracle who could see each customer's true
$\tau_i$ and email only those it pays for would realise \eur{\nbZeroPolOracle}. That money is not
in the average at all --- no ATE-level rule, however perfectly identified, can reach it --- and it
is exactly the prize \Cref{ch:up} competes for. The decision this chapter fixes is \emph{whether
to send}; the decision it cannot reach is \emph{whom to send to}.

\begin{remark}[The memo to the CMO]
The dashboard says the email pays for itself; it does not. Most of the gap is \emph{who we
emailed}, not what the email did: our historical targeting picked loyal, engaged customers who
would have spent more anyway (\eur{\nbZeroSelBias} of selection bias), and on whom the email also
happens to work better than average (\eur{\nbZeroTargetingGap} of ATT-above-ATE). What to do, in
order: \textbf{randomize} when you can, because a holdout kills selection bias by design; when you
cannot, \textbf{adjust} for the confounders you measured and \textbf{stress-test} the ones you did
not; and note that the real money is not in the average at all.
\end{remark}

% =====================================================================================
\section{Takeaways}
\label{sec:fnd:take}

\begin{enumerate}[leftmargin=*, itemsep=.45em]
  \item \textbf{Half of every unit's table is missing, and that is the whole problem.} Each
        customer has two potential outcomes and reveals one (\cref{tab:fnd:oracle}); causal
        inference imputes the other half under stated assumptions \citep{holland1986, ding2018}.
        The assumptions, not the arithmetic, are where the argument is won.

  \item \textbf{The naive difference is an effect plus a claim about who was chosen.}
        \Cref{eq:fnd:decomp} splits the dashboard's \eur{\nbZeroNaiveObs} into an ATT of
        \eur{\nbZeroAtt} plus \eur{\nbZeroSelBias} of selection bias --- the emailed customers'
        head start. Two distinct errors, neither noise.

  \item \textbf{Randomization is worth more than any estimator in this book.} A coin flip makes
        the selection term exactly zero and performs the graph surgery of \cref{fig:fnd:surgery}
        physically. The confounded estimate sits \nbZeroObsSeDist{} standard errors from the truth
        against the randomized \nbZeroRandSeDist{}; more data only makes a confounded estimate more
        precisely wrong.

  \item \textbf{The graph decides what to control for, and two of the three configurations say
        ``don't''.} Confounders are adjusted for; mediators and colliders are not. Conditioning on
        \emph{response} --- a common effect of the email and of spend --- manufactures
        \eur{\nbZeroColliderResp} per responder on a campaign whose true effect is exactly zero.
        Post-treatment variables stay out of the adjustment set, and M-bias (\cref{eq:fnd:mbias})
        shows even a \emph{pre}-treatment variable can be a trap.

  \item \textbf{Every adjusted rung works, and the standard error is part of the estimator.} On
        confounded data with a planted \eur{\nbZeroTrueDag}, regression, IPW and AIPW all recover
        it while the naive rung returns \eur{\nbZeroLadNaive} (\cref{tab:fnd:ladder}). Each demands
        a different covariance --- Welch, HC1, bootstrap --- and an interval on the wrong one is a
        confident lie.

  \item \textbf{No estimator upgrade fixes missing data.} With an unmeasured confounder, all four
        rungs fail \emph{together} and \emph{silently}: on a clean diagnostic sheet, AIPW returns
        \eur{\nbZeroHidPanelEst} inside a tight interval that is \eur{\nbZeroHidPanelBias} wrong.
        Double robustness insures against a wrong model of the features you \emph{have}, never
        against a feature you do \emph{not} have. \textbf{Estimation is sharpened by better methods;
        identification is not.}

  \item \textbf{The failure is real, and it is expensive.} Acting on the confounded number burns
        \eur{\nbZeroPolBurn} per \nbZeroNCustomers{} contacts, and on LaLonde it inverts the sign
        of a real effect. A diagnostic says \emph{where} an estimate is unsupported, not what the
        supported estimate would have been.
\end{enumerate}

\notebooknote{%
  This chapter is \code{notebooks/00\_foundations.ipynb}, \S1--\S6 and \S8--\S9; \Cref{ch:pvp} is
  its \S7. The simulators are \code{cmp.dgp.uplift\_customers} (\cref{eq:fnd:uplift}) and
  \code{cmp.dgp.dag\_control\_demo} (\crefrange{eq:fnd:dag_t}{eq:fnd:dag_y}); their constants are
  \cref{tab:fnd:params} and Appendix A. The classical estimators and their covariance choices are
  \code{cmp.classical} and \code{cmp.estimators}; the seeds are fixed in the notebook. Every number
  in this chapter comes from the full run (\code{CMP\_FAST=0}), with the LaLonde section enabled
  (\code{CMP\_REAL=1}, which fetches the Dehejia--Wahba extract of the NSW data from a public URL),
  emitted by \code{cmp.report} and injected here as a macro. The one exception is the
  \usd{\nbZeroLlBenchmark} experimental benchmark, which is cited from \citet{dehejia1999} rather
  than estimated, and is flagged as such where it is emitted.}

% =====================================================================================
\clearpage
\section*{Appendix 1.A\quad Formal machinery}
\label{app:fnd}
\addcontentsline{toc}{section}{Appendix 1.A\quad Formal machinery}

The body of this chapter states each idea in plain language, gives the euro result, and points
here. This appendix carries the formal version of each --- the definitions, the derivations, the
standard-error justifications, and the balance diagnostic --- every one with the worked example
the body refers to. Nothing here is needed to follow the chapter's decisions; it is the reference
a reader consults when a plain-language statement is not enough. (The planted DGP constants
themselves live in the book's global Appendix~A, alongside \cref{tab:fnd:params}.)

\subsection*{1.A.1\quad Potential outcomes, consistency, and the fundamental problem}

Fix a customer $i$. Before anything is done, that customer carries \emph{two} numbers, not one:
$Y_i(1)$, the spend if emailed, and $Y_i(0)$, the spend if not. These are the \textbf{potential
outcomes} \citep{neyman1923, rubin1974}. Both are properties of the customer; they exist before
the marketer decides anything, and the decision merely selects which one becomes visible:
%
\begin{equation}
  Y_i \;=\; T_i\,Y_i(1) \;+\; (1 - T_i)\,Y_i(0).
  \label{eq:fnd:consistency}
\end{equation}
%
\Cref{eq:fnd:consistency} is called \emph{consistency}: what you observe is the potential outcome
belonging to the treatment actually received. The other one is not merely unmeasured. It is
\textbf{unmeasurable} --- the customer was emailed, so the world in which she was not is gone.

The \textbf{individual causal effect} is the contrast $\tau_i = Y_i(1) - Y_i(0)$: what the email
did \emph{to that customer}. \Cref{eq:fnd:consistency} makes it permanently invisible, because
one of its two terms is always missing. This is the \emph{fundamental problem of causal
inference} \citep{holland1986}, and its most useful reading is the one \citet{ding2018} make
explicit: \textbf{causal inference is a missing-data problem}. Half of every customer's row was
never filled in, and every method in this book is a disciplined way of imputing it.

\Cref{tab:fnd:oracle} shows the missing half. It takes six real customers from the file and
prints \emph{both} potential outcomes --- possible only because this world is simulated, and the
simulator knows the branch it did not run. The daggered entries are the counterfactuals: in any
real customer file, those cells are blank.

\input{build/tables/nb00_oracle}

\begin{intuition}{Why one row is never enough}
Look down the \textbf{Emailed} column of \cref{tab:fnd:oracle} and cover, with a finger, every
daggered cell. What remains is what a real analyst has. For every emailed customer you can see
what she spent \emph{with} the email and never what she would have spent without it; for every
un-emailed customer, the reverse. No amount of cleverness recovers an individual's missing cell.
What can be recovered --- and only under assumptions, and only sometimes --- is an \emph{average}
over many customers. That is why this book estimates averages, and why the next subsection
insists on saying \emph{which} average.
\end{intuition}

\subsection*{1.A.2\quad The estimands: ATE, ATT, CATE (and LATE)}

Because individual effects are unavailable, decisions run on averages --- and \emph{which}
average is a choice that must be made before a method is chosen, because the three answer
different commercial questions about the same campaign:
%
\begin{align}
  \text{ATE} &\;=\; \E\big[Y(1) - Y(0)\big]
     &&\text{the effect on a randomly chosen customer,} \label{eq:fnd:ate}\\
  \text{ATT} &\;=\; \E\big[Y(1) - Y(0) \,\big|\, T = 1\big]
     &&\text{the effect on those actually emailed,} \label{eq:fnd:att}\\
  \tau(x) &\;=\; \E\big[Y(1) - Y(0) \,\big|\, X = x\big]
     &&\text{the CATE: the effect for customers like } x. \label{eq:fnd:cate}
\end{align}
%
The ATE prices \emph{roll it out to everyone?} The ATT audits \emph{what did our past sends
earn?} The CATE decides \emph{who next?} They are not interchangeable, and the difference is
commercial rather than terminological. \Cref{fig:fnd:tau} makes the point in one picture: the ATE
of \eur{\nbZeroTrueAte} sits \emph{below} the \eur{\nbZeroCost} cost, so an average-level rule
says \emph{email no one} --- and yet \pc{\nbZeroShareAboveCost} of customers individually clear
the cost bar, while \pc{\nbZeroShareNegative} are actively \emph{harmed} by the email. One
number, opposite per-customer instructions. Recovering $\tau(x)$ is the subject of \Cref{ch:up}.

A fourth average, the \textbf{LATE}, appears later in the book: the effect on the customers an
\emph{instrument} moves --- those who act on an encouragement but would not have acted without
it. It cannot be defined honestly here, because it needs instruments, and instruments need a
chapter. \Cref{ch:iv} gives it one; until then the word is not used.

\subsection*{1.A.3\quad The naive difference and its standard error}

The naive difference of means \cref{eq:fnd:naive} is computable from any customer file and needs
no assumptions to \emph{compute}; whether it means anything causal is the whole question. Every
estimate also carries a \textbf{standard error}: the typical distance between the estimate and the
number it is estimating, caused purely by the luck of which customers landed in the sample. For
\cref{eq:fnd:naive} it is
%
\begin{equation}
  \widehat{\se}\big(\hat\tau_{\text{naive}}\big)
    \;=\; \sqrt{\frac{s_1^{2}}{n_1} + \frac{s_0^{2}}{n_0}},
  \label{eq:fnd:welchse}
\end{equation}
%
with $s_t^2$ and $n_t$ the sample variance and the count in arm $t$. This is the \emph{Welch}
form \citep{welch1947}: it does not assume the two arms are equally variable. A standard error is
a statement about \emph{noise}, and about nothing else. It shrinks like $1/\sqrt n$. It has no
opinion whatever about whether the right customers were compared, which is why a confounded
estimate with a small standard error is not a good estimate --- it is a precisely wrong one, and
\cref{sec:fnd:valid} measures the difference.

\subsection*{1.A.4\quad Selection bias: the decomposition, derived}

Take the naive difference, apply consistency (\cref{eq:fnd:consistency}), and add and subtract the
one quantity nobody observes --- the mean untreated outcome \emph{among the treated},
$\E[Y(0) \mid T{=}1]$. The three-term identity \cref{eq:fnd:decomp} of \cref{sec:fnd:vocab}
follows at once. Two of its quantities deserve to be read slowly. $\E[Y(0) \mid T{=}1]$ is the
treated group's own baseline: what the emailed customers would have spent had they not been
emailed --- purely counterfactual, and not identified from the treated arm by any amount of data.
Its twin $\E[Y(0) \mid T{=}0]$ is the controls' baseline, and it \emph{is} observed. The naive
difference substitutes the second for the first, and \textbf{selection bias is exactly that
substitution error}.

\Cref{tab:fnd:oracle} lets us watch the identity hold, because the simulator knows every $Y(0)$.
Its six customers are drawn from the \emph{extremes} of the file --- the most engaged in the
emailed arm, the least engaged in the un-emailed one --- so the arithmetic is a dramatised version
of the full sample's, displayed at a scale where it is legible. Among the table's three emailed
customers the mean of $Y(0)$ is \eur{\nbZeroOrBaseTreated}; among its three un-emailed customers
it is \eur{\nbZeroOrBaseControl}. The emailed group's head start is therefore
$\eur{\nbZeroOrBaseTreated} - \eur{\nbZeroOrBaseControl} = \eur{\nbZeroOrSelBias}$ --- before any
email is sent. The observed means are \eur{\nbZeroOrTreatedMean} and \eur{\nbZeroOrControlMean}, so
the naive difference is \eur{\nbZeroOrNaive}, while the true average effect on those three emailed
customers is only \eur{\nbZeroOrAtt}. And \cref{eq:fnd:decomp} is exact:
%
\[
  \underbrace{\eur{\nbZeroOrNaive}}_{\text{naive}}
  \;=\; \underbrace{\eur{\nbZeroOrAtt}}_{\text{ATT}}
  \;+\; \underbrace{\eur{\nbZeroOrSelBias}}_{\text{selection bias}} .
\]
%
The overstatement is not an approximation, a bias term or a residual. It is the wrong baseline.
The same decomposition on all \nbZeroNCustomers{} customers is less theatrical and exactly as
damning; \cref{sec:fnd:vocab} reports it.

\subsection*{1.A.5\quad Graphs: $d$-separation, forks, chains, colliders, and the backdoor}

A \textbf{directed acyclic graph} (DAG) draws each variable as a node and each direct cause as an
arrow: $A \rightarrow B$ means ``$A$ is a direct cause of $B$''. \emph{Acyclic} means no variable
causes itself through any loop. A DAG is not a picture of the data and it is not recovered from the
data --- it is a \emph{claim about the world}, defended with domain knowledge, and it must be
committed to before the data are touched \citep{greenland1999}.

A \textbf{path} between $T$ and $Y$ is any trail of arrows connecting them, \emph{ignoring the
arrow directions}. Association flows along paths. A path is either \textbf{open} (association
flows) or \textbf{blocked} (it does not), and --- this is the whole trick --- conditioning on a
variable changes which. Three configurations exhaust the cases, and the rule for each is different
(\cref{fig:fnd:dags} draws all three):
%
\begin{description}[leftmargin=*, style=nextline]
  \item[Fork (a common cause): $T \leftarrow L \rightarrow Y$.]
    Loyalty $L$ raises both the chance of being emailed and the spend. The path is \textbf{open by
    default}, and it carries association that is not causal. Conditioning on $L$ \textbf{blocks} it.
  \item[Chain (a mediator): $T \rightarrow M \rightarrow Y$.]
    The email causes an open ($M$), which causes spend. The path is \textbf{open by default}, and
    it is \emph{part of the effect}. Conditioning on $M$ blocks it --- and thereby deletes part of
    the very thing being measured.
  \item[Collider (a common effect): $T \rightarrow C \leftarrow Y$.]
    Whether the customer responded, $C$, is caused by the email \emph{and} by how much she was
    going to spend. A path through a collider is \textbf{blocked by default}, and conditioning on
    $C$ \textbf{opens} it, manufacturing association that was never there.
\end{description}
%
The unifying statement is $d$-separation: \emph{a path is blocked if it passes through a fork or
chain that you condition on, or through a collider that you do not.} The same operation ---
conditioning --- has opposite effects depending on the structure, which is precisely why the
instinct to ``control for everything we have'' fixes the first case and breaks the other two.

A \textbf{backdoor path} is a path from $T$ to $Y$ that begins with an arrow pointing \emph{into}
$T$ --- as in $T \leftarrow L \rightarrow Y$. It is a route by which $T$ and $Y$ become associated
\emph{without} $T$ causing $Y$: a back door into the outcome. Confounding \emph{is} an open
backdoor path, and the job of adjustment is to close every one of them.

\subsection*{1.A.6\quad Adjustment and the g-formula}

The verb ``adjust'' is used constantly and is almost never defined. To \textbf{adjust} for $X$ ---
equivalently, to \emph{condition on} $X$, or to \emph{control for} $X$ --- means: compare treated
and untreated units \emph{only within groups that share the same value of $X$}, and then average
those within-group comparisons over the population. In practice it is done in one of three
interchangeable ways: \emph{stratify} (split into bins of $X$ and average the bins), \emph{regress}
(put $X$ in the model), or \emph{reweight} (make the arms' $X$-mixes match). Formally it is the
\textbf{adjustment formula}, also called the g-formula \citep{robins1992, hernan2020}:
%
\begin{equation}
  \text{ATE} \;=\; \sum_x
     \Big\{\ \E\big[Y \mid T{=}1,\, X{=}x\big] - \E\big[Y \mid T{=}0,\, X{=}x\big]\ \Big\}\,
     \Prob(X = x).
  \label{eq:fnd:gformula}
\end{equation}
%
Read it as an instruction, not a formula. The inner bracket is the conditioning: \emph{compare
like with like}. The outer sum is the averaging: \emph{over the population's mix of $X$}. The
naive difference of \cref{eq:fnd:naive} does the second part wrong --- it averages over the mix of
$X$ found \emph{inside each arm}, and the arms have different mixes. That single substitution is
the whole of selection bias, written in a second notation. \Cref{sec:fnd:ident} executes it by
hand on the one-confounder world (\cref{tab:fnd:strata}); \cref{sec:fnd:classical} executes it
four ways.

\subsection*{1.A.7\quad The estimator ladder: the standard error each rung demands}

The standard-error column of \cref{tab:fnd:ladder} is not bookkeeping. An interval built on the
wrong standard error is worse than no interval, because it is a confident lie --- and the four
rungs demand four different answers:

\begin{itemize}[leftmargin=*]
  \item \textbf{Naive.} The \emph{Welch} standard error of \cref{eq:fnd:welchse}
        \citep{welch1947}, which does not assume the two arms have equal variances. It costs
        almost nothing when they happen to be equal, and it is the only one of the two
        candidates still valid when they are not --- and which world one is in is not announced
        in advance.
  \item \textbf{Regression.} An \textbf{HC1} heteroskedasticity-robust standard error
        \citep{white1980}: insurance against an error variance that moves with $X$. It is worth
        being exact about what that insurance costs, because \cref{eq:fnd:dag_y} is
        \emph{homoskedastic by construction} --- its noise term does not depend on $L$ --- so the
        premium ought to be nothing, and ``ought to be'' is not a measurement. Refitting the
        identical model with an iid covariance gives a standard error of \eur{\nbZeroOlsSeIid}
        against HC1's \eur{\nbZeroOlsSeHc}: a gap of \pc{\nbZeroOlsSeGapPct}. That is the honest
        argument for a robust default, and it is not the one usually offered. HC1 is not reported
        here because the errors are heteroskedastic; it is reported because the premium is
        \emph{measurable}, it is negligible when the errors are well behaved, and real spend data
        --- right-skewed, zero-inflated and lumpy --- rarely is.
  \item \textbf{IPW.} No closed form survives the fact that the propensity is itself
        \emph{estimated}, so the interval is a percentile \textbf{bootstrap}
        \citep{efron1979, efron1993}: resample customers with replacement, refit the propensity
        \emph{and} the estimator, and take percentiles of the re-estimates. Refitting the
        propensity inside the loop is the point --- a bootstrap that treated $\hat e$ as fixed
        would price only half the uncertainty.
  \item \textbf{AIPW.} A cross-fitted influence-function estimate --- the \emph{influence
        function} being each customer's first-order contribution to the estimate --- with a
        bootstrap interval. It is a \emph{frequentist} estimator and its interval is a confidence
        interval, a distinction \Cref{ch:pvp} is entirely about.
\end{itemize}

\subsection*{1.A.8\quad The balance diagnostic: standardized mean difference}

The \emph{standardized mean difference} (SMD) of a feature is the gap in its mean between the arms,
measured in pooled standard deviations:
%
\begin{equation}
  \text{SMD} \;=\;
  \frac{\bar X_1 - \bar X_0}{\sqrt{\tfrac12\,(s_1^{2} + s_0^{2})}}\,.
  \label{eq:fnd:smd}
\end{equation}
%
Dividing by the spread is what makes it comparable across features measured in different units,
and the conventional bar is $|\text{SMD}| < 0.1$: below that, the arms are treated as balanced on
that feature. For the loyalty feature in the raw email data, \cref{eq:fnd:smd} gives
$\text{SMD} = \nbZeroSmdLoyalty$ --- far above the bar, and confounding made visible before any
weighting. \Cref{sec:fnd:valid} reports it before and after weighting; \cref{sec:fnd:real} reads
it on real data, where the worst imbalance is far larger.
