% =====================================================================================
% A note on the numbers — the book's one methodological commitment about itself.
%
% The single honest exception (LaLonde's $1,800 experimental benchmark, cited rather than
% computed) is stated here AND flagged in place in Chapter 1. It is itself emitted as a
% macro, so even the cited number is not retyped.
% =====================================================================================
\chapter*{A note on the numbers}
\addcontentsline{toc}{chapter}{A note on the numbers}
\markboth{A note on the numbers}{A note on the numbers}

\textbf{Every number in this book was injected from an executed notebook. None was retyped.}

The claim is mechanical rather than aspirational, and it is worth one page because it changes what
a reader is entitled to assume. Each analysis in this book lives in a runnable notebook, and each
notebook \emph{emits} its results --- scalars, tables and figures --- as it computes them. The
book then \emph{pulls them in}: a scalar becomes a macro, a table becomes a typeset
\code{booktabs} float, a figure becomes a vector page-sized PDF with the caption that was written
beside the computation that produced it. The prose cites the macro. So the sentence a reader meets
in a chapter and the number the notebook printed cannot disagree, because they are the same
object.

Two properties follow, and the second is the one that matters.

\emph{A stale number is impossible.} Re-run a notebook with a different seed, a longer chain or a
corrected model, rebuild, and every figure in the prose that depends on it moves. There is no
second copy of the result to fall out of date, because the prose never held a copy.

\emph{A missing number is a compilation error.} If a chapter cites a result the notebooks did not
emit, the typesetter halts and the book refuses to produce a PDF. This is deliberate and it is not
a nuisance: a hole is louder than a stale number, and the failure mode this whole apparatus exists
to prevent --- a paragraph that quietly describes an output the code stopped producing a year ago
--- fails silently by construction unless something is built to make it fail loudly.

The discipline extends to places where it is inconvenient. The epigraph that opens each chapter
carries real numbers, and they are macros. The caption under each figure carries real numbers, and
they were authored in the notebook, next to the computation, and travel with the figure. The
summary tables were not laid out by hand at all; they are the notebook's own data frames, rendered.

Three classes of literal do appear in the source, and they are literals precisely because they are
\emph{not} results. The constants of each data-generating model --- the coefficients this book
plants in its simulated worlds, collected in \cref{app:dgp} --- are definitions of the world, not
measurements of it. The decision thresholds ($\Prob > 0.9$ for \textsc{go}, $< 0.5$ for
\textsc{no-go}) and the interval level (90\,\%) are stated conventions, argued with in the text
rather than estimated from data. And the priors are model statements, defended in the section that
declares them.

\section*{The one exception, stated}

There is exactly one number in this book that was \emph{not} computed by the code that surrounds
it, and it deserves to be named on its own page rather than discovered in a footnote.

\Cref{ch:foundations} closes on \citet{lalonde1986}'s job-training data, where a randomized
evaluation and an observational comparison exist side by side --- which is the only configuration
in which a claim about adjustment can be graded on real data at all. The randomized benchmark ---
the experimental effect of training on 1978 earnings, about \usd{\nbZeroLlBenchmark} --- is
\textbf{cited from the published literature, not estimated here}. It is not recomputed from the
file the notebook analyses, which carries the trainees and the non-experimental comparison group
--- the observational arm, whose whole purpose is to be graded against the experiment. Everything
else in that section is computed --- the naive comparison, the adjusted estimates, the overlap
diagnostic, and the
\usd{\nbZeroLlSwing} the estimate moves under adjustment. The benchmark that grades them is
borrowed, and it is flagged as borrowed in the notebook, in the chapter, and here.

Even that number is not retyped. It is emitted as a macro like every other, carrying a note that
records its provenance as a citation. The exception is to the \emph{computation}, not to the
injection.
